Writing Trash Bin

### 2024-08-03 paper.qmd
finished writing "show me the data" section

Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA. However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation. It is unlikely that the methods discussed thus far are able to achieve this.

@@

It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by providing *more* information. Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates. Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021]. Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value. So is this the best uncertainty visualisation? And if so, why?

## Where is all this uncertainty coming from?
The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation. The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.

### variance is not uncertainty
This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with. Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate. Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake. We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty. The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves. This confusion is common and there are many papers that spend a great deal of time clarifying the difference. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. @Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.

You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.

Some authors simply confuse variance for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.

### Goal/task dependence in each stage
Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.

### Problems with uncertainty having no definition
The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

Unfortunately "the amalgamations of the impacts of violations to the assumption of the uniformity of nature" is a rather broad concept that does not easily roll off the tongue. Regardless, this definition of uncertainty does seem to align with what the literature considers to be included in the uncertainty umbrella. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision. These works are only concerned with quantifying the final impact of uncertainty on our statistics. That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred. Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity. These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.

Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication. Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.

## The distribution problem

## Show me the data
If we want to show uncertainty for the sake of

If you have uncertainty information, the best approach to create an "uncertainty visualisation" might be to incorperate that information into the visualisation to attempt to replicate the original sample.

@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size

@@

@@

>>>>

Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA. However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation. It is unlikely that the methods discussed thus far are able to achieve this.

@@

It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by simulating *more* information. The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots. Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates. Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021]. Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value. So is this the best uncertainty visualisation? And if so, why?

## Show me the data
The pixel map is not the best uncertainty visualisation, but it is trying to *imitate* the best uncertainty visualisation. The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.  As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". It's not a definition that rolls off the tongue, but we can work with it.

Thankfully, this definition of uncertainty aligns nicely with all the concepts that are included in the uncertainty umbrella. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision. These works are only concerned with quantifying the final impact of uncertainty on our statistics. That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred. Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity. These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.

What this means is that we have two types of uncertainty, but one is more "processed" than the other. Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic. The disconnect between these two expressions of uncertainty creates a huge problem for authors trying to visualise it. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect. The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty. There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify. Some authors don't believe that it is even possible to quantify all the assumption violations. @Spiegelhalter2017 mentioned that combining the uncertainties that appear at each stage of an analysis into a single "uncertainty" value is near impossible.

The second reason to leave uncertainty as a potential violation of our assumptions, is that we might not know the final statistic we are seeking to calculate. This is the case for visualisations made for EDA, and a large number of developments in EDA visualisation have been in displaying these difficult to quantify violations. For example, @Tierney2023 builds upon the tidy data principles to allow users to handle missing values. This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
Therefore the best uncertainty visualisation is not am "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data, that gives us a good idea of the limitations of our model.

This does not mean that visualizing raw data instead of implementing sampling techniques will always prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. While it is not always possible, it should always be considered as an effective uncertainty visualisation when the raw data is available.

@@

### Goal/task dependence in each stage
Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.  @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.

@@

@moritz2017trust did some trust EDA stuff? check the paper

### 2024-08-03 writingdumpster.qmd
finished writing "show me the data" section

A final method to perform signal suppression is simply to visualise the data if it is available and relevant to the uncertainty distribution. An example of this is shown in @fig-census for racial distributions spatially in America, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, (e.g. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out. This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. However it is simple but effective option that seems to be largely overlooked by the uncertainty visualisation community.

@@

>>>>

A final method to perform signal suppression is simply to visualise the data if it is available and relevant to the uncertainty distribution. An example of this is shown in @fig-census for racial distributions spatially in America, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, (e.g. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.

@@

# Variance is not uncertainty
Some authors simply confuse variance for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.

This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with. Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate. Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake. We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty. The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves. This confusion is common and there are many papers that spend a great deal of time clarifying the difference. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. @Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
